Mapping out the punishment strategy space

### 2023-11-24 manuscript.Rmd
finished manuscript first draft

---
title             : "Mapping the punishment strategy space"
shorttitle        : "Punishment strategies"

@@

abstract: |
 x

keywords          : "x"
wordcount         : "xxxx words"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \usepackage{xcolor}
  - \usepackage{makecell}
  - \renewcommand\theadfont{}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
  - \nolinenumbers
  - \note{\raggedright * Correspondence concerning this article should be addressed to Scott Claessens, Level 2, Building 302, 23 Symonds Street, Auckland, New Zealand. \text{E-mail:} \href{scott.claessens@gmail.com}{\nolinkurl{scott.claessens@gmail.com}} \par This working paper has not yet been peer-reviewed.}
appendix          : "appendix.Rmd"
csl               : nature.csl
---

@@

Humans cooperate on a scale unparalleled in the animal kingdom. One mechanism thought to sustain this level of cooperation is costly punishment, whereby individuals harm others at a personal cost[@CluttonBrock1995], ostensibly encouraging cooperative behaviour from the target (or bystanders[@DosSantos2011;@DosSantos2013;@Raihani2015]) in the future. Punishment therefore offers a route to maintaining or increasing cooperation by changing the payoff structure of social interactions such that it no longer pays to cheat or exploit social partners[@CluttonBrock1995;@Ostrom1990].

In humans, many studies of punishment have been carried out in laboratory settings using economic games[@Balliet2011;@Balliet2013;@Chaudhuri2011;@Dreber2008;@Fehr2000;@Fehr2002;@Henrich2006;@Nikiforakis2008;@Raihani2012a;@Raihani2019]. In these games, participants are usually given a sum of money that they can use to invest in collective action or to help others. Alternatively, participants can 'cheat' by keeping the money for themselves or by exploiting the contributions of others. Punishment is introduced into such games by giving participants the option to pay a small 'fee' to impose a greater 'fine' on their co-players. Several lines of experimental evidence indicate that people use this punishment option[@Henrich2006], that they enjoy punishing[@DeQuervain2004], and that they frequently, though not always[@Herrmann2008], punish cheating or exploitative co-players[@Fehr2000;@Fehr2002].

Evidence from these experiments suggests that the threat of costly punishment plays an important role in promoting human cooperation. People tend to cooperate more in games where punishment is possible compared to those where it is not[@Balliet2011;@Balliet2013;@Raihani2019]. The effect that the threat of punishment has on cooperation is also evident in the higher contributions typically observed in the Ultimatum Game (where punishment is possible) compared to the structurally-similar Dictator Game (where it is not)[@Camerer2003]. This typical cooperation-enhancing effect of punishment has also been observed across societies[@Balliet2013], leading some to suggest that costly punishment has played a key role in the cultural evolution of cooperation in humans[@Bowles2013;@Boyd2003;@Chudek2011;@Henrich2017].

Nevertheless, it remains unclear whether individuals playing economic games use punishment as a behaviour-change tool to enforce cooperation or as a means to achieve other ends. Some have argued that punishment is primarily used to shape future behaviour, either to deter personal harm[@Delton2017;@Dreber2008;@DosSantos2013] or to uphold normative standards of cooperative behaviour[@Boyd2003;@Chudek2011;@Fehr2018;@Mathew2011;@Mathew2014;@Richerson2016]. But while the _threat_ of punishment can have a cooperation-enhancing effect, the _enactment_ of this punishment seldom changes the future behaviour of deviant targets[@Raihani2019]. This calls into question whether punishment primarily operates as a behaviour-change tool or whether it is used to achieve other goals.

Beyond behaviour-shaping concerns, there are a host of other reasons that people may want to punish in economic games. Punishers might be motivated by a desire for retribution rather than deterrence, punishing in proportion to the amount of harm that was personally caused[@Carlsmith2002]. Punishment might be driven by concerns about relative payoffs, such as disadvantageous inequity aversion (i.e., avoiding having less than others[@Raihani2012b;@Raihani2019]) and/or general egalitarian preferences (i.e., wanting all participants to receive the same payoffs[@Dawes2007]). Such concerns about relative payoffs may be activated when participants earn less than cheats in economic games or when there are income disparities in these settings. People might also use punishment for competitive purposes, seeking advantageous inequity for themselves (i.e., having more than others) and/or improving their relative position[@Raihani2019].

Common economic game designs have been unable to tease apart these different motives for punishment because participants who interact with cheats in these games experience both losses _and_ lower relative payoffs. The typical 1:3 fee-to-fine ratio of punishment in economic games compounds this issue. With this setup, people can simultaneously use punishment to reciprocate losses, to deter others from cheating, and to reduce or reverse disparities in payoffs between themselves and targets. To add to this complexity, it is evident that people use punishment in seemingly disparate ways: punishing when no behaviour change is possible, such as in one-shot games[@Bone2015;@Henrich2006;@Raihani2012b;@Walker2004] or in games where the target never learns about the punishment[@Crockett2014]; punishing those who did not cheat or who over-contributed to collective action (antisocial punishment[@Herrmann2008;@Sylwester2013]); punishing in scenarios where they were not personally harmed (third-party punishment[@Fehr2004]); and punishing in scenarios where disparities in payoffs did not arise from participants' actions[@Dawes2007;@Johnson2009;@Fowler2005].

@@

This insight raises several underexplored questions. First, which punishment strategies are more frequent in human populations? Second, what traits predict adherence to a particular strategy? Previous work has reported that personality is related to cooperative behaviour[@Thielmann2020] and demographics, political ideology, and religiosity are related to punitive behaviour[@Hofmann2018], but no research has related these variables to specific punishment strategies. Third, do people have insight into their own punishment strategy? Previous work has argued that people are often unaware of the underlying function of their punitive behaviour, yet they feel compelled to enact it anyway[@Carlsmith2002;@Nisbett1977].

Here, we aim to delineate the many possible punishment strategies by asking whether people punish in a manner consistent with a specific strategy and, if so, what other characteristics (personality, social preferences, political orientation) predict the use of different punishment strategies. Table \@ref(tab:tableStrategies) summarises the potential functions for costly punishment in economic games and the behavioural strategies they predict. Note that we do not include reputational functions of punishment in this table, such as signalling trustworthiness[@Barclay2006;@Batistoni2022;@Jordan2016;@Jordan2017;@Raihani2015], because our focus is on punishment strategies in anonymous economic games without reputational incentives (but see ref[@Jordan2020]).

(ref:tableStrategiesCaption) xxx

```{r tableStrategies, echo=FALSE}
apa_table(
  tar_read(tableStrategies),
  escape = FALSE,
  landscape = TRUE,
  align = c("p{1.7cm}", "p{3.3cm}", rep("c", 12)),
  col_spanners = list(
    "\\thead{No DI 1 \\\\ 70-10}"   = 3:4,
    "\\thead{No DI 2 \\\\ 70-30}"   = 5:6,
    "\\thead{No DI 3 \\\\ (Computer) \\\\ 70-30}"   = 7:8,
    "\\thead{No DI 4 \\\\ (1:1 Fee-Fine) \\\\ 70-30}"   = 9:10,
    "\\thead{DI \\\\ 70-50}"        = 11:12,
    "\\thead{3PP \\\\ 70-70 [100]}" = 13:14
    ),
  caption = "(ref:tableStrategiesCaption)",
  font_size = "scriptsize"
)
```

Building on previous designs[@Bone2015;@Deutchman2021;@Marczyk2017;@Raihani2012b], we employ a suite of one-shot economic games where individuals are given the opportunity to punish targets at a personal cost. We carefully designed this suite of games to tease apart the proposed punishment strategies in Table 1, such that each strategy predicts a different pattern of behaviour across all the games. We use the resulting behavioural patterns to discern which punishment strategy participants are employing. We then combine these behavioural patterns with data on demographics, personality, political ideology, religiosity, and self-reported strategy usage.

@@

x

@@

x

>>>>

---
title             : "Why do people punish? Evidence for a range of strategic concerns"
shorttitle        : "Punishment strategies"

@@

abstract: |
 Costly punishment is thought to be one of the key mechanisms sustaining cooperation in humans. However, the motives for punitive behaviour remain unclear. Punishment is often assumed to be motivated by a desire to convert free riders into cooperators, but it is also consistent with a host of other functions, such as levelling payoffs or increasing one’s relative position. We used a suite of six economic games to tease apart the different motives for punishment. Across representative samples from the United Kingdom and the United States, we estimated the frequency of different punishment strategies in the population, finding that egalitarian motives for punishment are more common than behaviour-change motives. Moreover, different punishment strategies were differentially predicted by personality, social preferences, political ideology, and religious views. Self-reports of behaviour in the games suggested that people have some degree of insight into their punishment strategy. These findings highlight the multipurpose nature of human punishment.

keywords          : "punishment; cooperation; economic games"
wordcount         : "5185 words"
bibliography      : ["r-references.bib"]
floatsintext      : yes
figurelist        : no
figsintext        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
numbersections    : false
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
header-includes: 
  - \usepackage{array}
  - \usepackage{caption}
  - \usepackage{graphicx}
  - \usepackage{siunitx}
  - \usepackage[normalem]{ulem}
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{hhline}
  - \usepackage{calc}
  - \usepackage{tabularx}
  - \usepackage{threeparttable}
  - \usepackage{wrapfig}
  - \usepackage{adjustbox}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \usepackage{xcolor}
  - \usepackage{makecell}
  - \raggedbottom
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1,small}}
  - \captionsetup[figure]{font={stretch=1,small}}
  - \nolinenumbers
  - \note{\raggedright * Correspondence concerning this article should be addressed to Scott Claessens, Level 2, Building 302, 23 Symonds Street, Auckland, New Zealand. \text{E-mail:} \href{scott.claessens@gmail.com}{\nolinkurl{scott.claessens@gmail.com}} \par This working paper has not yet been peer-reviewed.}
appendix          : "appendix.Rmd"
csl               : nature.csl
---

```{r options, echo=FALSE}
opts_chunk$set(echo = FALSE)
```

@@

Humans cooperate on a scale that is unparalleled in the animal kingdom. One 
mechanism thought to sustain this level of cooperation is costly punishment, 
whereby individuals harm others at a personal cost[@CluttonBrock1995], 
ostensibly encouraging cooperative behaviour from the target (or bystanders [@DosSantos2011;@DosSantos2013;@Raihani2015]) in the future. Punishment 
therefore offers a route to maintaining or increasing cooperation by changing 
the payoff structure of social interactions such that it no longer pays to cheat 
or exploit social partners[@CluttonBrock1995;@Ostrom1990].

In humans, many studies of punishment have been carried out in laboratory 
settings using economic games[@Balliet2011;@Balliet2013;@Chaudhuri2011;
@Dreber2008;@Fehr2000;@Fehr2002;@Henrich2006;@Nikiforakis2008a;@Raihani2012a;
@Raihani2019]. In these games, participants are usually given a sum of money 
that they can use to invest in collective action or to help others. 
Alternatively, participants can 'cheat' by keeping the money for themselves or 
by exploiting the contributions of others. Punishment is introduced into such 
games by giving participants the option to pay a small 'fee' to impose a greater 
'fine' on their co-players. Several lines of experimental evidence indicate that 
people use this punishment option[@Henrich2006], that they enjoy 
punishing[@DeQuervain2004], and that they frequently, though not 
always[@Herrmann2008], punish cheating or exploitative 
co-players[@Fehr2000;@Fehr2002].

Evidence from these experiments suggests that the threat of costly punishment 
plays an important role in promoting human cooperation. People tend to cooperate 
more in games where punishment is possible compared to those where it is not
[@Balliet2011;@Balliet2013;@Raihani2019]. The effect that the threat of 
punishment has on cooperation is also evident in the higher contributions 
typically observed in the Ultimatum Game (where punishment is possible) compared 
to the structurally-similar Dictator Game (where it is not)[@Camerer2003]. This 
typical cooperation-enhancing effect of punishment has also been observed across 
societies[@Balliet2013], leading some to suggest that costly punishment has 
played a key role in the cultural evolution of cooperation in humans
[@Bowles2013;@Boyd2003;@Chudek2011;@Henrich2017].

Nevertheless, it remains unclear whether individuals playing economic games use
punishment as a behaviour-change tool to enforce cooperation or as a means to 
achieve other ends. Some have argued that punishment is primarily used to shape 
future behaviour, either to deter personal harm[@Delton2017;@Dreber2008;
@DosSantos2013] or to uphold normative standards of cooperative behaviour 
[@Boyd2003;@Chudek2011;@Fehr2018;@Mathew2011;@Mathew2014;@Richerson2016]. 
But while the _threat_ of punishment can have a cooperation-enhancing effect, 
the _enactment_ of this punishment does not consistently deter targets from 
cheating in the future[@Raihani2019]. This calls into question whether 
punishment primarily operates as a behaviour-change tool or whether it is used 
to achieve other goals.

Beyond behaviour-shaping concerns, there are a host of other reasons why people 
may want to punish in economic games. Punishers might be motivated by a desire 
for retribution rather than deterrence, punishing in proportion to the amount 
of harm that was personally caused[@Carlsmith2002]. Punishment might be driven 
by concerns about relative payoffs, such as disadvantageous inequity aversion 
(i.e., avoiding having less than others[@Raihani2012b;@Raihani2019]) and/or 
general egalitarian preferences (i.e., wanting all participants to receive the 
same payoffs[@Dawes2007]). Such concerns about relative payoffs may be activated 
when participants earn less than cheats in economic games or when there are 
income disparities in these settings. People might also use punishment for 
competitive purposes, seeking advantageous inequity for themselves (i.e., having 
more than others) and/or improving their relative position[@Raihani2019].

Common economic game designs have been unable to tease apart these different 
motives for punishment because participants who interact with cheats in these 
games experience both losses _and_ lower relative payoffs. The typical 1:3 
fee-fine ratio of punishment in economic games compounds this issue. With 
this setup, people can simultaneously use punishment to reciprocate losses, to 
deter others from cheating, and to reduce or reverse disparities in payoffs 
between themselves and targets. To add to this complexity, it is evident that 
people use punishment in seemingly disparate ways: punishing when no behaviour 
change is possible, such as in one-shot games[@Bone2015;@Henrich2006;
@Raihani2012b;@Walker2004], on the very last round of repeated games
[@Barclay2016], or in games where the target never learns about the 
punishment[@Crockett2014]; punishing those who did not cheat or who 
over-contributed to collective action (antisocial punishment
[@Herrmann2008;@Sylwester2013]); punishing in scenarios where they were not 
personally harmed (third-party punishment[@Fehr2004]); and punishing in 
scenarios where disparities in payoffs did not arise from participants' 
actions[@Dawes2007;@Johnson2009;@Fowler2005].

@@

This insight raises several underexplored questions. First, which punishment 
strategies are more frequent in human populations? Second, what traits predict 
adherence to a particular strategy? Previous work has reported that personality 
is related to cooperative behaviour[@Thielmann2020] and demographics, political 
ideology, and religiosity are related to punitive behaviour[@Hofmann2018], but 
no research has related these variables to specific punishment strategies. 
Third, do people have insight into their own punishment strategy? Previous work 
has argued that people are often unaware of the underlying function of their 
punitive behaviour, yet they feel compelled to enact it anyway
[@Carlsmith2002;@Nisbett1977].

Here, we aim to delineate nine possible punishment strategies by asking 
whether people punish in a manner consistent with a specific strategy and, if 
so, what other characteristics (personality, social preferences, political 
orientation) predict the use of different punishment strategies. Table 
\@ref(tab:tableStrategies) summarises the potential functions for costly 
punishment in the economic games that we considered, and the behavioural 
strategies they predict. Note that Table \@ref(tab:tableStrategies) is not 
an exhaustive list of all possible punishment strategies: we do not include 
reputational functions of punishment in this table, such as signalling 
trustworthiness[@Barclay2006;@Batistoni2022;@Jordan2016;@Jordan2017;@Raihani2015], 
because our focus is on punishment strategies in anonymous economic games without 
reputational incentives (but see ref[@Jordan2020]).

Building on previous designs[@Bone2015;@Deutchman2021;@Marczyk2017;
@Raihani2012b], we employ a suite of one-shot economic games where individuals 
are given the opportunity to punish targets at a personal cost (Figure 
\@ref(fig:plotGames)). In each game, targets either steal from another 
individual or do nothing. Representative samples of participants from the United 
Kingdom (*n* = `r sum(tar_read(dExc)$Country == "United Kingdom")`) and the 
United States (*n* = `r sum(tar_read(dExc)$Country == "United States")`) 
completed all six games on the online platform Prolific. We carefully designed 
the suite of games to tease apart the proposed punishment strategies in Table 
\@ref(tab:tableStrategies), such that each strategy predicts a different pattern 
of behaviour across all the games (see Methods for more detail about the six 
games). We use the resulting behavioural patterns to discern which punishment 
strategy participants are employing. We then combine these behavioural patterns 
with data on demographics, personality, social preferences, political ideology,
religiosity, and self-reported strategy usage.

(ref:plotGamesCaption) *Visual summary of the six economic games.* In all
games, Player 2 either steals £0.20 from Player 1 (the focal player) or does 
nothing. Player 1 is then given the option to punish by paying a certain amount 
of money to remove money from Player 2 (this money is destroyed). The six games
are variants on this general setup, creating situations where (A) Player 2 is
still worse off by stealing, (B) Player 2 creates equality by stealing, (C) the
computer "decides" whether Player 2 steals, (D) the fee-fine ratio is 1:1, (E)
Player 2 is better off by stealing, and (F) Player 2 steals instead from a 
third-party.

```{r plotGames, out.width="100%", fig.cap="(ref:plotGamesCaption)"}
include_graphics("images/games.png")
```

(ref:tableStrategiesCaption) \scriptsize Summary of the different functions for 
punishment and the behavioural strategies they predict. *Games A-F are 
the games employed in the current study (see Methods for more details). In each 
of the six games, participants are given the opportunity to punish players who 
"steal" and those who do not, meaning that participants make twelve punishment 
decisions in total. Each behavioural strategy implies a unique pattern of 
punishment across all decisions. Green ticks reflect decisions to punish, red 
crosses reflect decisions to not punish. In column headers, payoffs at the first 
stage (above) and the second stage (below) are denoted as P1-P2 (or P2-P3 [P1] 
for Game F) where participants take the role of P1 and P2 is the target of 
punishment. AI = advantageous inequity, DI = disadvantageous inequity.*

```{r tableStrategies}
apa_table(
  tar_read(tableStrategies),
  escape = FALSE,
  landscape = TRUE,
  align = c("p{1.7cm}", "p{3.3cm}", rep("c", 12)),
  col_spanners = list(
    "\\thead{\\scriptsize Game A \\\\ (AI) \\\\ 70-10}"   = 3:4,
    "\\thead{\\scriptsize Game B \\\\ (Equal) \\\\ 70-30}"   = 5:6,
    "\\thead{\\scriptsize Game C \\\\ (Computer) \\\\ 70-30}"   = 7:8,
    "\\thead{\\scriptsize Game D \\\\ (1:1 Fee-Fine) \\\\ 70-30}"   = 9:10,
    "\\thead{\\scriptsize Game E \\\\ (DI) \\\\ 70-50}"        = 11:12,
    "\\thead{\\scriptsize Game F \\\\ (Third-Party) \\\\ 70-70 [100]}" = 13:14
    ),
  caption = "(ref:tableStrategiesCaption)",
  font_size = "scriptsize"
)
```

@@

```{r MLMs}
mlm1 <- coef(summary(tar_read(mlm1)))
mlm2 <- coef(summary(tar_read(mlm2)))
```

The overall pattern of punitive behaviour in the six economic games was in line 
with previous research and very similar across both countries (Figure 
\@ref(fig:plotDecisions)). Participants were generally more likely to punish 
targets who stole from another individual compared to targets who did not steal 
(multilevel logistic regression; _b_ = 
`r printnum(mlm1["otherTook","Estimate"])`, standard error = 
`r printnum(mlm1["otherTook","Std. Error"])`, _p_ 
`r printp(mlm1["otherTook","Pr(>|z|)"])`). Participants were also more likely 
to punish when targets' stealing behaviour generated inequalities, 
specifically in Games E and F (_b_ = 
`r printnum(mlm2["DI","Estimate"])`, SE = `r printnum(mlm2["DI","Std. Error"])`, 
_p_ `r printp(mlm2["DI","Pr(>|z|)"])`).

(ref:plotDecisionsCaption) *Overall pattern of punitive behaviour across all six 
economic games, split by country.* AI = advantageous inequity, DI = 
disadvantageous inequity.

```{r plotDecisions, fig.width=8, fig.height=5, fig.cap="(ref:plotDecisionsCaption)"}
tar_read(plotDecisions)
```

We classified participants into a particular strategy if their behaviour across 
all twelve decisions matched our behavioural predictions shown in Table 
\@ref(tab:tableStrategies) exactly. Table \@ref(tab:tableStrategyCounts) shows 
the proportion of participants following each strategy, with N/A used to 
represent participants who did not fit exactly into any particular strategy type.
Overall, `r printnum(mean(tar_read(dExc)$strategy != "N/A")*100, digits = 0)`%
of our participants could be classified exactly into one of the strategies. The
most common strategy in both countries was to never punish across any of the
games. The next most common strategies were those that care about
minimising payoff differences (avoid disadvantageous inequity, egalitarian). 
Less common were the behaviour-shaping strategies (deterrent, norm-enforcing),
the retributive strategy, and the competitive strategies (seek AI, competitive). 
Although participants often punished targets who did not steal 
in the six games (Figure \@ref(fig:plotDecisions)), no participants followed the 
antisocial strategy by exclusively punishing targets who did not steal across 
_all_ games.

(ref:tableStrategyCountsCaption) Counts and proportions of participants 
following each punishment strategy exactly, split by country. N/A implies that
participants were unable to be classified exactly into any of the punishment
strategies.

```{r tableStrategyCounts}
apa_table(
  tar_read(tableStrategyCounts),
  escape = FALSE,
  align = c("l", rep("c", 4)),
  col_spanners = list(
    "\\thead{United Kingdom \\\\ (N = 1014)}" = 2:3,
    "\\thead{United States \\\\ (N = 996)}" = 4:5
    ),
  caption = "(ref:tableStrategyCountsCaption)"
)
```

To further investigate the strategies that participants were following, we 
examined the most common patterns of punitive behaviour across all twelve 
decisions. Supplementary Table \@ref(tab:tablePatterns) shows the proportion of 
participants following the 25 most common behavioural patterns, including, 
where appropriate, the predetermined strategies from Table 
\@ref(tab:tableStrategies). In both countries, a common pattern of behaviour 
not captured by any of the strategies was punishing only when the target stole
in the third-party game (Game F). Punishment in this game is consistent with an 
egalitarian motive, as stealing produces unequal outcomes, but third-party 
punishment here is also consistent with norm-enforcing and competitive motives 
(see Table \@ref(tab:tableStrategies)). Other common behavioural patterns not 
captured by our strategies included punishing whenever the target stole across 
all games and always punishing in every game irrespective of the targets' 
behaviour.

While it is useful to look at exact patterns of behaviour, participants may not 
have implemented their chosen punishment strategy with exact precision.
In reality, strategies may have been implemented probabilistically for each 
punishment decision. There is also the possibility of implementation errors, 
whereby participants occasionally "slip up" and make decisions that are 
incongruent with a particular strategy. This may explain why some participants 
were unable to be classified exactly into a single punishment strategy.

To deal with this complexity and include all observed data in our frequency 
estimates, we fitted a Bayesian latent state model to the data. This model 
assumes that the nine strategies in Table \@ref(tab:tableStrategies) (plus a 
"random choice" strategy that chooses randomly for each decision) are the only 
latent strategies and that these are instantiated into observed behaviour 
according to the logic in Table \@ref(tab:tableStrategies) with some 
probability of implementation error (i.e., an intention to punish is implemented 
as non-punishment and vice versa). Averaging over all strategies and 
incorporating the possibility of implementation errors, the model estimates the 
probability of participants following any particular strategy, given the 
observed data.

The posterior estimates from the model are presented in Figure 
\@ref(fig:plotModel1b). The posterior probabilities for each strategy did not
differ between the two countries. In both countries, the never punish strategy
had the highest probability, followed by the egalitarian strategy. The
norm-enforcing and seek AI strategies were the next most likely, with higher
posterior estimates than the competitive and antisocial strategies. None of the
other strategies differed in their posterior estimates. The same general pattern
emerged when we analysed the full dataset without pre-registered exclusions 
(Supplementary Figure \@ref(fig:plotModel1a)).

(ref:plotModel1bCaption) *Posterior estimates of the probabilities of following
different punishment strategies from the Bayesian latent state model.* The model 
assumes an implementation error rate of 5%. Points represent posterior medians, 
line ranges represent 50% and 95% credible intervals.

```{r plotModel1b, fig.width=7, fig.height=5, fig.cap="(ref:plotModel1bCaption)", warning=FALSE}
tar_read(plot1.2)
```

Next, we explored which traits predicted adherence to different punishment 
strategies. To answer this question, we included variables capturing 
demographics, personality, social preferences, political views, and religious 
views as predictors in our Bayesian latent state model. We included each 
variable in a separate model, predicting all ten punishment strategies 
(the nine from Table 1, plus the 'random choice' strategy) simultaneously.

```{r sesSlope}
ses <- tar_read(post2.2_SES)$beta[,2,10]
```

Demographic variables tended to be unrelated to strategy usage: age and gender
did not predict adherence to a particular punishment strategy (Supplementary 
Figures \@ref(fig:plotAllDems2) and \@ref(fig:plotAllDems1)). In the United 
States, the never punish strategy was slightly more common among participants 
lower in socio-economic status (median posterior slope = 
`r printnum(median(ses))`, 95% CI [`r printnum(quantile(ses, 0.025))` 
`r printnum(quantile(ses, 0.975))`]) but this effect was small.

Conversely, personality and social preferences were linked to variation in 
punishment strategies. When including the Big-6 personality dimensions and 
Social Value Orientation (SVO) in the model, we found associations with the 
egalitarian, never punish, and random choice strategies (Figure
\@ref(fig:plotAllPers2)). Participants higher in SVO were more likely to follow 
the egalitarian and the never punish strategies, while those with lower SVO 
scores were more likely to enact the random choice strategy. The personality 
dimensions of honesty-humility and openness to experience were both positively 
associated with following the never punish strategy, while extraversion 
negatively predicted this strategy. The effects were mostly similar across
countries, but occasionally differed: for example, in the United States,
but not in the United Kingdom, honesty-humility was positively associated with 
following the egalitarian strategy and negatively associated with following 
the random choice strategy. Overall, the same pattern of results emerged when 
analysing the full dataset without exclusions (Supplementary Figure 
\@ref(fig:plotAllPers1)).

(ref:plotAllPers2Caption) *Posterior slopes from Bayesian latent state models 
including Big-6 personality dimensions and Social Value Orientation.* Each row 
represents a separate model. All models assume an implementation error rate of 
5%. Points represent posterior medians, line ranges represent 95% credible 
intervals.

```{r plotAllPers2, fig.width=7.5, fig.height=3.5, fig.cap="(ref:plotAllPers2Caption)", warning=FALSE}
tar_read(plotAllPers2)
```

Political and religious variables were also associated with punishment strategy 
(Figure \@ref(fig:plotAllPolRel2)). These effects tended to be more pronounced 
in the United States. Controlling for Social Dominance Orientation, American 
participants higher in Right Wing Authoritarianism were more likely to follow 
the strategies avoiding disadvantageous inequity and seeking advantageous 
inequity. Participants who stated that they would like to "bring those below 
them [on the socio-economic status ladder] up a peg" were more likely to follow 
the egalitarian strategy, while American participants higher in Social Dominance 
Orientation, Right Wing Authoritarianism, and believing that God controls events 
in the world were less likely to follow the egalitarian strategy. In general, 
religious and conservative participants were less likely to follow the never 
punish strategy. This general pattern of results was replicated with the full 
dataset (Supplementary Figure \@ref(fig:plotAllPolRel1)).

(ref:plotAllPolRel2Caption) *Posterior slopes from Bayesian latent state models 
including political ideology, views about social inequality, and religiosity.* 
Each row represents a separate model aside from Social Dominance Orientation and 
Right Wing Authoritarianism, which control for one another within the same 
model. All models assume an implementation error rate of 5%. Points represent 
posterior medians, line ranges represent 95% credible intervals.

```{r plotAllPolRel2, fig.width=7.5, fig.height=3.5, fig.cap="(ref:plotAllPolRel2Caption)", warning=FALSE}
tar_read(plotAllPolRel2)
```

Finally, we asked whether participants had insight into their own punishment 
strategy. In other words, could participants self-report the strategy that they 
were following during the games? To answer this question, we included 
participants' responses to post-game questions about their strategy as 
predictors in the model. As before, each predictor was included in a separate 
model, predicting all ten strategies simultaneously.

In general, we found that self-reported strategy usage was positively associated
with the behavioural strategy that participants employed (see Supplementary 
Figures \@ref(fig:plotSliders1) and \@ref(fig:plotSliders2) for the distribution 
of responses to self-report questions). Figure \@ref(fig:plotAllSliders2) shows 
the relationships between self-report questions and the different punishment 
strategies, highlighting the combinations where the question matched the 
behavioural strategy. We found positive relationships between the self-report 
questions and strategy usage for the norm-enforcing, egalitarian, seek 
advantageous inequity, never punish, and random choice  strategies. The 95% 
credible intervals for other estimates included zero, though these estimates 
often trended in a positive direction. The same pattern of results was found 
when analysing the full dataset without exclusions  (Supplementary Figure 
\@ref(fig:plotAllSliders1)).

(ref:plotAllSliders2Caption) *Posterior slopes from models including 
self-reported strategy usage.* Each row represents a separate model. All models 
assume an implementation error rate of 5%. Highlighted estimates represent 
combinations where the self-report question matched the behavioural strategy. 
Points represent posterior medians, line ranges represent 95% credible 
intervals.

```{r plotAllSliders2, fig.width=7.5, fig.height=4, fig.cap="(ref:plotAllSliders2Caption)", warning=FALSE}
tar_read(plotAllSliders2)
```

@@

Using a suite of economic games measuring punishment in different situations, 
we have shown that punishment does not serve just one function, but instead is 
a flexible tool that can be and is used for different purposes[@Raihani2019]. 
Punishment is more akin to a swiss army knife than a hammer, used by some to 
enforce norms of cooperation and by others to reduce or even create inequality 
between individuals. We found that people's punishment strategy can, to some 
extent, be predicted by individual differences in personality, social 
preferences, and political and religious views. Moreover, contrary to the view 
that people are often unable to articulate the reasons for their punitive 
behaviour[@Carlsmith2002], people seem to have some degree of insight into the 
strategy they are using. Despite small differences, these general patterns 
replicated in samples from both the United Kingdom and the United States.

Among the punitive strategies, the most common were particularly sensitive to 
inequality in payoffs, either from a self-referential perspective (i.e., avoid 
disadvantageous inequity) or more generally (i.e., egalitarian). This is in 
line with previous studies which have highlighted inequity aversion as an 
important motivation for punishment in economic games 
[@Dawes2007;@Raihani2012b]. Personality and social preference variables mapped 
onto these strategies in expected ways. Traits associated with other-regarding 
concern, such as SVO and honesty-humility, predicted following the egalitarian 
strategy, whereas religious and conservative individuals were less likely to 
follow this strategy, especially in the United States. Moreover, participants 
following the egalitarian strategy were able to self-report this strategy, 
though the same was not true for the avoid disadvantageous inequity strategy.

Behaviour shaping strategies, such as deterrence and norm-enforcement, were less 
common than strategies sensitive to inequality in our set of games. This was 
reflected both in participants’ elicited punishment behaviour (Figure 
\@ref(fig:plotModel1b)) and in their self-reports of their own strategy 
(Supplementary Figures \@ref(fig:plotSliders1) and \@ref(fig:plotSliders2)). 
Regarding the predictors of these strategies, we found that demographic, 
personality, political, and religious variables tended to be unrelated to 
deterrent and norm-enforcing punishment strategies. We also found that 
participants had insight into the norm-enforcing strategy, but not the deterrent 
strategy. This finding is in line with previous research showing that people 
struggle to accurately report the deterrent motivations for their punitive 
behaviour[@Carlsmith2002].

Other punitive strategies were less common in our dataset, but some were more 
prevalent than others. For example, participants were more likely to use 
punishment to seek advantageous inequity than to exclusively harm those who did 
not steal (i.e., antisocial punishment). The existence of the "seek AI" strategy 
in our dataset supports the claim that punishment can also be used as a tool to 
increase one’s own relative position[@Raihani2019]. While generally rare, we 
found that this motive for punishment was more common among authoritarian 
participants, at least in the United States, potentially providing an 
explanation for why peer punishment has been found to be more common among 
conservatives in previous work[@Hofmann2018]. Moreover, the fact that no 
participants in our sample punished non-stealing across all games suggests that 
antisocial punishment does not function to harm cooperators specifically, as has 
been previously suggested[@Herrmann2008]. Instead, antisocial punishment appears 
to be motivated by improving one's relative position in general, which is in 
line with work showing that antisocial punishment disappears with a 1:1 fee-fine 
ratio[@Sylwester2013].

The fact that people use punishment for many different reasons poses problems 
for the way that punishment is operationalised in classic behavioural economic 
game studies. In these studies, a common assumption is that participants will 
punish to change the behaviour of cheats. But in reality, people may be choosing 
the punishment option to achieve a variety of different goals. The targets of 
punishment in these studies are likely well aware that punishment could be 
levied for these different reasons and this knowledge may impact their 
responses. For example, if cheating targets interpret punishment as serving a 
competitive motive, it may elicit retaliation rather than encourage 
cooperation[@Dreber2008;@Nikiforakis2008b;@Raihani2019]. This might help to 
explain the mixed findings in the field as to whether punishment actually 
motivates cheating targets to cooperate in the future[@Raihani2019].
